Estimating initial state and quark-gluon plasma medium properties\\ using a hybrid model with nucleon substructure\\ calibrated to p-Pb and Pb-Pb collisions at \texorpdfstring{$\mathbf{\sqrts=5.02}$}{}~TeV

### 2018-07-11 paper/substructure.tex
Start writing Bayesian section

\usepackage[labelfont={color=theblue}, justification=raggedright, labelsep=quad]{caption}

@@

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\sqrts}{\sqrt{s_{NN}}}
\newcommand{\sigmaf}{\sigma_\mathrm{fluct}}
\newcommand{\taufs}{\tau_\mathrm{fs}}
\newcommand{\n}{n_\mathrm{c}}
\newcommand{\R}{R_\mathrm{c}}
\newcommand{\w}{w_\mathrm{c}}
\newcommand{\dmin}{d_\mathrm{min}}
\newcommand{\T}{\tilde{T}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\X}{\mathcal{X}_\mathrm{struct}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

@@

\section{nuclear collision model}
\label{sec:model}

@@

\begin{figure}
  \begin{tikzpicture}
    % spherical proton
    \draw[dashed, xshift=-2.5cm] (0,0) circle (1cm);
    % three partons
    \draw[dashed] (0,0) circle (1cm);
    \foreach \theta in {0, 120, 240}{
      \draw ({\theta}:.5) circle (.4cm);
    }
    % ten partons
    \draw[dashed, xshift=2.5cm] (0,0) circle (1cm);
    \foreach \theta/\radius in {
      0/0.6, 40/0.5, 90/0.6, 130/0.5, 180/0.6, 220/0.6,
      260/0.5, 320/0.6, 300/0.2
    }{
      \draw[xshift=2.5cm] ({\theta}:\radius) circle (.3cm);
    }
  \end{tikzpicture}
  \caption{
    \label{fig:substructure}
    Schematic of plausible proton shapes.
    The sketch on the left shows a spherically symmetric proton (dashed line), while the middle and right illustrations depict a fluctuating proton with three and nine constituents respectively (solid lines).
  }
\end{figure}

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node {\includegraphics{thickness}};
    \node[label=above:{10 fm}] (a) at (0, 1.1) {};
    \node (b) at (0, -1.1) {};
    \draw [<->, semithick] (a) to (b);
  \end{tikzpicture}
  \caption{
    \label{fig:thickness} Effect of nucleon substructure on the nuclear thickness function $T(x, y) \equiv \int dz\, \rho(x, y, z)$ of a $^{208}\mathrm{Pb}$ nucleus.
    The nucleus on the left has Gaussian nucleons of width $0.88$~fm, while the nucleus on the right has composite nucleons, each containing six Gaussian constituents of width $0.41$~fm.}
\end{figure}

@@

\fig[t]{viscosity_dof}{
Degrees of freedom in the temperature dependent shear and bulk viscosity parametrizations.
Lines are chosen for illustrative purposes only and do not represent all possible variability.
For instance, $\eta/s$ could have a large slope and negative curvature, or $\zeta/s$ could have a large max and narrow width, neither of which are depicted above.
}

@@

\section{parameter estimation}
\label{sec:calibration}

The nuclear collision model constructed in Sec.~\ref{model} includes a number of free parameters $\x$ which describe the initial state, pre-equilibrium dynamics, and hydrodynamic medium.
We assume that these parameters are embedded in a realistic model such that each parameter is meaningful, but we concede that we know very little about their true values in nature.

\begin{table}
  \caption{
    \label{tab:design}
    Input parameter ranges for the initial condition and hydrodynamic models.
  }
  \begin{ruledtabular}
  \begin{tabular}{lll}
    Parameter         & Description                        & Range             \\
    \paddedhline
    Norm              & Normalization factor                 & 9--28           \\
    $p$               & Entropy deposition parameter         & $-1$ to $+1$    \\
    $\sigmaf$         & Nucleon fluctuation std.\ dev.\      & 0--2            \\
    $\w$              & Constituent sampling width           & 0.4--1.2 fm     \\
    $\n$              & Number of constituents               & 1--9            \\
    $\X$              & Nucleon structure parameter          & 0--1            \\
    $\dmin$           & Minimum inter-nucleon distance       & 0--1.7 fm       \\
    $\taufs$          & Free-streaming time                  & 0.1--1.5 fm/$c$ \\
    $\eta/s$ min      & Minimum value of $\eta/s$ (at $T_c$) & 0--0.2          \\
    $\eta/s$ slope    & Slope of $\eta/s$ above $T_c$        & 0--8 GeV$^{-1}$ \\
    $\eta/s$ crv      & Curvature of $\eta/s$ above $T_c$    & $-1$ to $+1$    \\
    $\zeta/s$ max     & Maximum value of $\zeta/s$           & 0--0.1          \\
    $\zeta/s$ width   & Width of $\zeta/s$ peak              & 0--0.1 GeV      \\
    $\zeta/s$ $T_0$   & Temperature of $\zeta/s$ maximum     & 150--200 MeV    \\
    $T_\text{switch}$ & Switching/particlization temp.       & 135--165 MeV    \\
  \end{tabular}
  \end{ruledtabular}
\end{table}

\section{results}
\label{sec:results}

@@

\fig*[t]{observables_design}{
Hybrid model calculations compared to experimental data for $p$-Pb collisions (left column) and Pb-Pb collisions (right column) at $\sqrts=5.02$~TeV.
Colored lines are model calculations (no emulator) for each of the 500 design points.
Points with error bars are experimental data from ALICE and CMS \note, with statistical and systematic errors added in quadrature.
}

\fig*[t]{observables_posterior}{
Calibrated emulator predictions compared to experimental data for $p$-Pb collisions (left column) and Pb-Pb collisions (right column) at $\sqrts=5.02$~TeV.
Colored lines are emulator predictions for $n=100$ different parameter sets randomly sampled from the Bayesian posterior.
Points with error bars are experimental data from ALICE and CMS \note, with statistical and systematic errors added in quadrature.
}

\begin{figure*}[p]
  \makebox[\textwidth]{
    \includegraphics[width=.95\paperwidth]{posterior}
  }
  \captionsetup{
    margin={.3\textwidth,0pt},
    justification=raggedleft,
    skip=-.75\textheight
  }
  \caption{Bayesian posterior distribution of the model input parameters. The diagonal panels show the marginalized distributions of individual\\ model parameters, while the off-diagonal panels show the joint\\ distributions for pairs of model parameters, visualizing\\ their correlations. The marginalized distribution\\ medians are annotated along the diagonal.}
  \label{fig:posterior}
  \vspace{.5\textheight}
\end{figure*}

>>>>

\usepackage[font=small, justification=raggedright, labelsep=quad]{caption}

@@

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\sqrts}{\sqrt{s_{NN}}}
\newcommand{\sigmaf}{\sigma_\mathrm{fluct}}
\newcommand{\taufs}{\tau_\mathrm{fs}}
\newcommand{\n}{n_\mathrm{c}}
\newcommand{\R}{R_\mathrm{c}}
\newcommand{\w}{w_\mathrm{c}}
\newcommand{\dmin}{d_\mathrm{min}}
\newcommand{\T}{\tilde{T}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\yexp}{\mathbf{y_\mathrm{exp}}}
\newcommand{\X}{\mathcal{X}_\mathrm{struct}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

@@

\section{Nuclear collision model}
\label{sec:model}

@@

\begin{figure}
  \begin{tikzpicture}
    % spherical proton
    \draw[dashed, xshift=-2.5cm] (0,0) circle (1cm);
    % three partons
    \draw[dashed] (0,0) circle (1cm);
    \foreach \theta in {0, 120, 240}{
      \draw ({\theta}:.5) circle (.4cm);
    }
    % ten partons
    \draw[dashed, xshift=2.5cm] (0,0) circle (1cm);
    \foreach \theta/\radius in {
      0/0.6, 40/0.5, 90/0.6, 130/0.5, 180/0.6, 220/0.6,
      260/0.5, 320/0.6, 300/0.2
    }{
      \draw[xshift=2.5cm] ({\theta}:\radius) circle (.3cm);
    }
  \end{tikzpicture}
  \caption{Schematic of plausible proton shapes. The sketch on the left shows a spherically symmetric proton (dashed line), while the middle and right illustrations depict a fluctuating proton with three and nine constituents respectively (solid lines).}
  \label{fig:substructure}
\end{figure}

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node {\includegraphics{thickness}};
    \node[label=above:{10 fm}] (a) at (0, 1.1) {};
    \node (b) at (0, -1.1) {};
    \draw [<->, semithick] (a) to (b);
  \end{tikzpicture}
  \caption{Effect of nucleon substructure on the nuclear thickness function $T(x, y) \equiv \int dz\, \rho(x, y, z)$ of a $^{208}\mathrm{Pb}$ nucleus. The nucleus on the left has Gaussian nucleons of width $0.88$~fm, while the nucleus on the right has composite nucleons, each containing six Gaussian constituents of width $0.41$~fm.}
  \label{fig:thickness}
\end{figure}

@@

\fig[t]{viscosity_dof}{Degrees of freedom in the temperature dependent shear and bulk viscosity parametrizations. Lines are chosen for illustrative purposes only and do not represent all possible variability. For instance, $\eta/s$ could have a large slope and negative curvature, or $\zeta/s$ could have a large max and narrow width, neither of which are depicted above.
}

@@

\section{Parameter estimation}
\label{sec:calibration}

The nuclear collision model constructed in Sec.~\ref{sec:model} includes a number of free parameters $\x$ which describe the initial state, pre-equilibrium dynamics, and hydrodynamic medium.
Given values for the parameters $\x$, the model may be used to predict a vector of experimental observables $\y$.
For example, $\y$ might be a vector of fixed length which includes charged particle yields in different centrality bins.
The physics model thus represents a vector-valued function $f(\x) = \y$
which maps the parameter values $\x$ to the observable state $\y$.

The goal of this work is to estimate the true model parameters $\x_\star$, given some evidence that our model ${f: X \mapsto Y}$ describes the experimental data $\yexp$ for the simulated observables $\y$.
According to Bayes' theorem, the probability for the true model parameters $\x_\star$ is
\begin{equation}
  \label{eq:bayes}
  P(\x_\star|f, \yexp) \propto P(f, \yexp|\x_\star) P(\x_\star).
\end{equation}
The left-side of this expression is the \emph{posterior}: the probability that $\x_\star$ is the correct value given evidence from the model $f: X \mapsto Y$ and the experimental data $\yexp$.

On the right-side, the \emph{likelihood} function $P(f, \yexp | \x_\star)$ is the compatibility of the experimental evidence $\yexp$ with the hypothetical model parameters $\x_\star$, and the \emph{prior} probability $P(\x_\star)$ is the estimate of the probability of $\x_\star$ in the absence of evidence $\yexp$.

\begin{equation}
  P(H1 | E, H2) \propto P(E | H1, H2)\, P(H1 | H2)
\end{equation}
There are three types of events,
\begin{itemize}
  \item $H1$: hypothesis that the model parameters $\x_\star$ are the true model parameters.
  \item $H2$: hypothesis that nuclear collisions are describable using the hybrid transport model formulated in this work.
  \item $E$: experimental evidence provided by data $\yexp$.
\end{itemize}

\subsection{Parameter design}

We create a scaffolding of the parameter space and distribute $n=500$ parameter points in our 15-dimensional parameter space using a Latin hypercube design \note.
The algorithm maximizes the

We assume a flat prior distribution for each model parameter $x \in \x$ described by the range of values shown in Tbl.~\ref{tab:design}.
The minimum and maximum values of the parameters are conservative to avoid clipping the calibrated posterior; for example, previous analyses of \trento\ model found $p \sim 0$, but we use the prior $-1 < p < 1$ to account for physics differences in the present model which might modify its posterior.

\subsection{Gaussian process emulator}

\begin{table}
  \caption{Input parameter ranges for the initial condition and hydrodynamic models.}
  \begin{ruledtabular}
  \begin{tabular}{lll}
    Parameter         & Description                        & Range             \\
    \paddedhline
    Norm              & Normalization factor                 & 9--28           \\
    $p$               & Entropy deposition parameter         & $-1$ to $+1$    \\
    $\sigmaf$         & Nucleon fluctuation std.\ dev.\      & 0--2            \\
    $\w$              & Constituent sampling width           & 0.4--1.2 fm     \\
    $\n$              & Number of constituents               & 1--9            \\
    $\X$              & Nucleon structure parameter          & 0--1            \\
    $\dmin$           & Minimum inter-nucleon distance       & 0--1.7 fm       \\
    $\taufs$          & Free-streaming time                  & 0.1--1.5 fm/$c$ \\
    $\eta/s$ min      & Minimum value of $\eta/s$ (at $T_c$) & 0--0.2          \\
    $\eta/s$ slope    & Slope of $\eta/s$ above $T_c$        & 0--8 GeV$^{-1}$ \\
    $\eta/s$ crv      & Curvature of $\eta/s$ above $T_c$    & $-1$ to $+1$    \\
    $\zeta/s$ max     & Maximum value of $\zeta/s$           & 0--0.1          \\
    $\zeta/s$ width   & Width of $\zeta/s$ peak              & 0--0.1 GeV      \\
    $\zeta/s$ $T_0$   & Temperature of $\zeta/s$ maximum     & 150--200 MeV    \\
    $T_\text{switch}$ & Switching/particlization temp.       & 135--165 MeV    \\
  \end{tabular}
  \end{ruledtabular}
  \label{tab:design}
\end{table}

\section{Results}
\label{sec:results}

@@

\begin{figure}[h]
  \makebox[\textwidth]{
    \begin{minipage}[t]{\paperwidth}
      \includegraphics[width=.95\paperwidth]{posterior}
      \captionsetup{
        width={.83\textwidth},
        justification=raggedright,
      }
      \caption{Bayesian posterior distribution of the model input parameters. The diagonal panels show the marginalized distributions of individual model parameters, while off-diagonal panels show the joint distributions for pairs of model parameters, visualizing their correlations. The marginalized distribution medians and 90\% credible intervals are annotated along the diagonal.}
    \end{minipage}
  }
  \label{fig:posterior}
\end{figure}

\begin{figure*}[p]
  \includegraphics{observables_pbpb}
  \caption{Simulated observables compared to experimental data for Pb-Pb collisions at $\sqrts=5.02$~TeV. Top row: explicit model calculations (no emulator) for each of the 500 design points; bottom row: emulator predictions for $n=100$ random samples drawn from the posterior. Points with error bars are experimental data from ALICE with statistical and systematic errors added in quadrature \note.}
  \vspace{5ex}
  \includegraphics{observables_ppb}
  \caption{Same as above but for $p$-Pb collisions at $\sqrts=5.02$~TeV. Note that multiplicity bins are used for mean $p_T$ and flow cumulant observables to match the bins used by experiment. Experimental data are from ALICE and CMS \note.}
\end{figure*}
